\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-14 paper.qmd
Rewrite the section "Preprocessing steps"

Consider the high-dimensional data a rectangular matrix $X$, where $X = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}$, with $n$ observations in $p$ dimensions. The objective is to discover a low-dimensional projection $Y = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}$, represented as an $n$ × $d$ matrix, where $d \ll p$. The reduction process seeks to remove noise from the original data set while retaining essential information.

@@

To tackle the complexities and noise in high-dimensional data, we apply PCA as a pre-processing step [@article67, @article68, @article69]. This step helps in noise reduction by identifying principal components that represent directions of maximum variance, capturing essential patterns in the data.

### Constructing the 2D model

@@

  After computing hexagonal grid configurations, the 2D emdeddings are binned into hexagon cells. As a result of hexagonal binning , the bin centroids ($C_k^{(2)} \equiv (C_{kx}, C_{ky}$)) (see @fig-meth (Step 2)) are obtained.

@@

Creating the model in the high-D space involves a two-step process. Initially, we identify observations within each hexagonal bin. After this, we calculate the average of the high-dimensional data within each hex bin. The resulting values represent the averaged high-dimensional data points ($C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p}$)). Essentially, these averaged high-dimensional data points serve as the representation of the hex bin centroids in the high-D space.

>>>>

Consider the high-dimensional data a rectangular matrix $X$, where $X = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top$, with $n$ observations in $p$ dimensions. The objective is to discover a low-dimensional projection $Y = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top$, represented as an $n$ × $d$ matrix, where $d \ll p$. The reduction process seeks to remove noise from the original data set while retaining essential information.

@@

In order to reduce the computational complexity associated with performing NLDR techniques to high-D data, and as a initial step of noise reduction of high-D data, PCA [@article67, @article68, @article69] is applied as a preprocessing step. Subsequently, the identified principal components, representing directions of maximum variance, are used to construct the model.

### Constructing the 2D model {#sec-construct2d}

@@

  After computing hexagonal grid configurations, the 2D emdeddings are binned into hexagon cells. As a result of hexagonal binning , the bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2}$) (see @fig-meth Step 2) are obtained.

@@

#### Lifting the triangular mesh points into high dimensions

**Step1: Cluster 2D points to hexagons**

Expanding upon the information regarding hexagonal binning discussed in Step 1 of @sec-construct2d, the primary objective in this step is to determine the 2D embedding points associated with each hexagon. As the initial process of hexagonal binning, the 2D points are clustered into their respective hexagonal bins. By mapping this information with the hexagonal bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$) that obtained in Step 2 of the 2D model building (see @sec-construct2d), we can find which 2D points are assigned to which data set (see @fig-wkhighD (a)).

**Step2: Cluster high dimensional points to hexagons**

Following the step 1, the main focus in this step to determine the corresponding high dimensional points for each hexagon. Every 2D embedding point serves as a projection of a data point belonging to high dimensional space. By using this mapping between the high dimensional data and their corresponding projections, the high dimensional points allocated to each hexagons are determined (see video linked in @fig-wkhighD).

**Step3: Compute the mean within hexagon**

Having identified the high-dimensional points associated with hexagons, the final step involves computing the mean within each hexagonal bin. This implies calculating the average of the high-dimensional data points located within each hex bin. These averaged high-dimensional data points, denoted as $C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})$, serve as the representative coordinates for the hex bin centroids within the expansive high-dimensional space (see video linked in @fig-wkhighD).

#### Lifting the 2D triangular mesh into high dimensions

Based on the insights gained in Step 3 of @sec-construct2d, where connected edges in the 2D triangular mesh were identified, this step involves lifting these connections into the high dimensions. Having the mappings of all 2D triangular mesh points into high dimensions, the points connected in 2D are also connected in high dimensional space (see video linked in @fig-wkhighD).

### 2024-01-14 paper.tex
Rewrite the section "Preprocessing steps"

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, boxrule=0pt, interior hidden, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

To tackle the complexities and noise in high-dimensional data, we apply
PCA as a pre-processing step \citep[\citet{article68},
\citet{article69}]{article67}. This step helps in noise reduction by
identifying principal components that represent directions of maximum
variance, capturing essential patterns in the data.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

In order to reduce the computational complexity associated with
performing NLDR techniques to high-D data, and as a initial step of
noise reduction of high-D data, PCA \citep[\citet{article68},
\citet{article69}]{article67} is applied as a preprocessing step.
Subsequently, the identified principal components, representing
directions of maximum variance, are used to construct the model.
